---
title: "Bar Plots with ggplot2"
author: "Tesfahun Tegene Boshe"
date: "06/11/2021"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Instructions

* Find interesting dataset and prepare short report (in R Markdown) which will consist:
  + short description of the dataset,
  + 4 barplots which will present interesting relationships between variables,
  + brief comments which describes obtained results.
  + Then, edit theme of the graphs and all scales of the graph and prepare publication-ready plots.

## Necessary packages

```{r}

requiredPackages = c("ggplot2","dplyr","tidyr")

for(i in requiredPackages){if(!require(i,character.only = TRUE)) install.packages(i)}
for(i in requiredPackages){if(!require(i,character.only = TRUE)) library(i,character.only = TRUE)}

```

## Dataset

For this task, we will analyse the English Premier League seasons 2000 - 2020. The source data is available at: https://www.kaggle.com/saife245/english-premier-league.

We will be using only the first 6 columns. The variables are explained below:

*Variable*      | *Description*
* X             |  number
* Date          |  Date of the match   
* HomeTeam      |  The team that played at home stadium
* AwayReam      |  The team that played away from home
* FTHG          |  Full time number of goals for the home team
* FTAG          |  Full time number of goals for the away team

```{r,echo=FALSE,eval = FALSE}
setwd(your_path)
```


```{r}

Data_PLeague <- read.csv2("Premier_league.csv", sep = ",")[,c(1:6)] # only the first 6 columns
head(Data_PLeague)

```

Let's add a column for the winner team and a column for total number of columns. The value should be "draw" in case of a draw. 
```{r}

Data_PLeague$Winner <- ifelse(Data_PLeague$FTHG>Data_PLeague$FTAG,Data_PLeague$HomeTeam, ifelse(Data_PLeague$FTHG<Data_PLeague$FTAG,Data_PLeague$AwayTeam,"Draw"))

Data_PLeague$GoalsScored <- factor(Data_PLeague$FTHG + Data_PLeague$FTAG) # define as factor
```

Let's convert the date from a character to date format. 
```{r}
Data_PLeague$Date <- as.Date(Data_PLeague$Date, "%d/%m/%y") # converts to date
Data_PLeague$Year <- format(Data_PLeague$Date, format = "%Y")
```

## Plots

### Plot1: Barplot of count of matches by total number of goals {.tabset}

#### Before edit
```{r}

p1 <- ggplot(data = Data_PLeague, aes(x = Data_PLeague$FTHG + Data_PLeague$FTAG)) + 
      geom_bar(stat = "count", color = "#3d195b", fill = "#3d195b") + 
      labs(x = "Number of goals",
           y = "Number of matches",
           title = "Count of the number of matches by number of goals")

p1
```

*Comments*

* Most Premier League matches experience 1-3 goals. This feeds to the general consensus that the Premier League is very defensive. 



#### After Edit
```{r}

p1 <- ggplot(data = Data_PLeague, aes(x = Data_PLeague$FTHG + Data_PLeague$FTAG)) + 
      geom_bar(stat = "count", color = "#7a0d32", fill = "#7a0d32") + 
      labs(x = "Number of goals",
           y = "Number of matches",
           title = "Count of the number of matches by number of goals") +
      theme_minimal() +
      theme(plot.title = element_text(hjust = 0.5),
            panel.background = element_rect(fill = "#d1d6ce", colour = "#d1d6ce", size = 1), #panel background
            plot.background = element_rect(colour = "#151414", size = 1) # plot background
            ) + 
       scale_x_discrete(limits = c(0:12))
        
      

p1
```

### Plot2: Number of wins by team {.tabset}

Data preparation
```{r}
winsByTeam <- Data_PLeague %>% 
  group_by(Winner) %>% summarize(n_wins = n())%>% 
  mutate(Winner2 = reorder(Winner, desc(n_wins))) %>% 
  slice_max(n_wins, n = 7) # top 6 plus draws

```

#### Before edit
```{r}

p2 <- ggplot(data = winsByTeam[-1,], aes(x = Winner2, y = n_wins)) + # exclude draws
      geom_col(color = "#7a0d32", fill = "#7a0d32") + # fill with premier league brand color
      ggtitle('Number of wins of the top-6 Premier League teams') +
      xlab('Team') +
      ylab('Number of wins') + 
      geom_text(aes(label = n_wins), vjust = -0.5, color = 'black',face = "bold", size = 3.5)

p2
```
*Comments*

* As of the season 2019/2020, Manchester United have had won the most games in the Premier League era followed by Chelsea. 
* Manchester City and Tottenham had won the same number of games since the start of the Premier League era until the end of 2019/2020. 

#### After edit

```{r}
p2 + theme_minimal() +
     theme(plot.title = element_text(hjust = 0.5),
          panel.background = element_rect(fill = "#d1d6ce", colour = "#d1d6ce", size = 1), #panel background
          plot.background = element_rect(colour = "#151414", size = 1) # plot background
        
          ) + 
     scale_y_continuous(breaks = seq(0, 600, by = 100), 
               limits = c(0, 600),
                ) 
```

### Plot3: Number of wins by Manchester United by year {.tabset}

#### Before edit
```{r}
Data_PLeague_MU <- Data_PLeague[Data_PLeague$Winner == "Man United" & Data_PLeague$Year >=2011 ,]# the last decade

p3 <- ggplot(data = Data_PLeague_MU, aes(x = Year)) + 
      geom_bar(stat = "count",aes(y = ..count..), fill = "#DA020E")+
      labs(x = "Year", 
           y = "Number of wins",
           title = "Manchester United number of wins per season")

p3

```
*Comments*

* 2012 registered the most wins for Man United. *Van Persie - Rooney* combination was the most lethal apparently. Of course defense wins games, as they say.   
* The team has been falling in quality since the season 2013. *Fergie's* departure must have had something to do with that. 

#### After edit
```{r}

p3 + theme_minimal() +
     theme(plot.title = element_text(hjust = 0.5),
          panel.background = element_rect(fill = "#d1d6ce", 
                                          colour = "#d1d6ce", 
                                          size = 1), #panel background
          plot.background = element_rect(colour = "#151414", 
                                         size = 1) # plot background
          ) + 
          scale_y_continuous(breaks = seq(0, 40, by = 5), 
                     limits = c(0, 40),
                      ) 
```

### Plot4: Number of matches drawn per year {.tabset}

I believe this metric indicates how competitive the league has been across years. More number of draws indicating less competitiveness. 

#### Before edit
```{r}
p4 <- ggplot(data = Data_PLeague[Data_PLeague$Winner=="Draw",],aes(x = Year)) + 
      geom_bar(stat = "count", color = "#3d195b", fill = "#3d195b") + 
      labs(x = "year", 
           y = "Draws in a season",
           title = "Number of matches drawn per year")

p4
```

*Comments*

* Generally, more and more matches ended at draw since the season 2003. Many boring matches :( 
* 2018 was the exception to that. Most matches ended with a winner. 

#### After edit
```{r}
p4 + 
  scale_x_discrete(limits = c('2000','2005','2010','2015','2020'))+ 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = .5),
        
        axis.title.x  = element_text(size = 12, hjust = 0.6), # labels of the x axis
        axis.title.y  = element_text(size = 12, hjust = 0.6))
  
```


### Plot5: Average Number of points per game {.tabset}

#### Before edit
Data preparation
```{r}
Data_PLeague$Points_HT <- ifelse(Data_PLeague$FTHG>Data_PLeague$FTAG,3,                                                ifelse(Data_PLeague$FTHG<Data_PLeague$FTAG,0,1))

points_H_team <- Data_PLeague %>% 
  group_by(HomeTeam) %>% summarize(avg_ppg = mean(Points_HT))%>% 
  mutate(HomeTeam2 = reorder(HomeTeam, desc(avg_ppg))) %>%
  slice_max(avg_ppg, n = 6)

```

plot

```{r}
p5 <- ggplot(data = points_H_team, aes(x = HomeTeam2, y = avg_ppg)) + 
      geom_col(color = "#3d195b", fill = "#3d195b")+
      geom_text(aes(label = round(avg_ppg,3)), vjust = -0.5, color = 'black', size = 3.5) + 
      labs(x = "Team", 
               y = "Points per game",
               title = "Points per home match")

p5
```
*Comments*

* Despite the recent fall in the form, Man United remains the most successful interms of points per game. Arsenal and Chelsea follow with small margin. 


#### After edit
```{r}
p5 + 
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
      panel.background = element_rect(fill = "#d1d6ce", 
                                      colour = "#d1d6ce", 
                                      size = 1), #panel background
      plot.background = element_rect(colour = "#151414", 
                                     size = 1) # plot background
      ) + 
   scale_y_continuous(breaks = seq(0, 3, by = 0.5), # create more space above the plots
                      limits = c(0, 3),
                  ) 
```

