---
title: "Homework 1/ Scatter Plots with ggplot2 package"
author: "Tesfahun Tegene Boshe"
date: "04/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Instructions

* Find interesting dataset and prepare short report (in R Markdown) which will consist:
- short description of the dataset,
- 3 scatterplots which will present interesting relationships between variables,
- brief comments which describes obtained results.

## Necessary packages

```{r}

# # install the libraries
# install.packages('ggplot2')
# install.packages('dplyr')
# install.packages('tidyr')
# install.packages('carData')

# load the libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(carData) # source for the data
```

## Dataset

For this task, Salaries dataset,The 2008-09 nine-month academic salary for Assistant Professors, Associate Professors and Professors in a college in the U.S  will be used.
The data were collected as part of the on-going effort of the college's administration to monitor salary differences between male and female faculty members.

*Variable*      | *Description*

* rank          | a factor with levels AssocProf AsstProf Prof
* discipline    | a factor with levels A (“theoretical” departments) or B (“applied” departments).
* yrs.since.phd | years since PhD.
* yrs.service   | years of service.
* sex           | a factor with levels Female Male
* salary        | nine-month salary, in dollars.

```{r}
data(Salary)
```

## plots

### Plot 1. Salary vs years of service for each sex

```{r}
ggplot(data=Salaries, aes(x=yrs.service,y = salary, color=sex))+
  geom_point() +
  labs(x = "Years of Service")+
  ggtitle("Salary vs Years of Service per sex") + 
  geom_smooth(method = "lm", color = "black")
```

comments:

* majority of the female professors are paid less than the market average
* there is a positive correlation between years of service and salary variables. 
* this is with the exception to the age group 40-60 years of service, where there is a general decrease in salary
* with years of experience

### Plot 2.1 Salary vs years since PHD for each rank 

```{r}
ggplot(data=Salaries, aes(x=yrs.since.phd,y = salary, color=rank))+
  geom_point() +
  labs(x = "Years since PHD")+
  ggtitle("Salary vs Years since PHD")
```


Comments

* Salaries increase with along the journey asstProf - AssociProf - Prof
* For each category, there is a positive correlation between salary and years since PHD

### Plot 2.2 Salary vs years since PHD for each discipline

```{r}

ggplot(data=Salaries, aes(x=yrs.since.phd,y = salary, color=discipline))+
  geom_point() +
  labs(x = "Years since PHD")+
  ggtitle("Salary vs Years since PHD")
```

Comments:

* applied departments pay more than theoretical departments

### plot 2.3 Salary vs years since PHD for each rank X discipline

```{r}
Salaries$Rank_Disc <- paste(Salaries$rank,"_", Salaries$discipline) 
# a new variable Rank_Disc is a cross product of rank and discipline

ggplot(data=Salaries, aes(x=yrs.since.phd,y = salary, color=Rank_Disc))+
  geom_point() +
  labs(x = "Years since PHD")+
  ggtitle("Salary vs Years since PHD")

```

Comments:

* Professors in applied departments are of the highest market value
* The properties in 2.1 and 2.2 are observed here too. 


### Plot 3. Bar plot of counts of individuals per sex per discipline

```{r}
ggplot(data=Salaries, aes(x=discipline, fill=sex))+
  geom_bar(stat="count", position="dodge") + #add position="dodge" to geom_bar to have the groups next to each other
  labs(x = "Discipline")+
  ggtitle("Sex distribution by discipline") +
  scale_fill_manual(values=c("red", "blue")) 

```

Comments:

* In both departments the majority of professors are male. 
* There are more professors in theoretical departments than in applied - depatments
  